---
title: "MOTU analyses"
author: "Eva Egelyng Sigsgaard"
date: "2023-09-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, message=FALSE}
library(dplyr)
library(vegan)
library(RColorBrewer)
library(ggplot2)
library(ggpubr)
library(ggrepel)
library(phyloseq)
library(hrbrthemes)
library(viridis)
library(e1071)
library(FSA)
library(ggVennDiagram)
library(ggvenn)
library(eulerr)
library(tidyverse)
library(ggforce)
library(car)
```

```{r load data}

## Get rarefied OTU table
otus1<-read.table("../../Tekstfiler/COI/COI_70/otu_rarefy_70.txt", header=T)

## Get tax table
taxa<-read.table("../../Tekstfiler/COI/COI_70/tax_rarefy_70.txt", sep="\t", row.names = 1, header=T)

##Get metadata
metadata<-read.table("../../Tekstfiler/COI/COI_70/metadata_rarefy_70.txt", sep="\t", header=T)

## Get more metadata
areas<-read.table("../../Tekstfiler/COI/Mads_metadata_230921_C19.txt", sep="\t", row.names=1, header=T, check.names=FALSE)

#Create extra metadata variables
metadata$sshc<-paste(metadata$substrate_type, metadata$season, metadata$habitat, metadata$cluster, sep="_")
##Including the number of field reps per sample
bio_reps<-metadata %>% group_by(sshc) %>% dplyr::summarise(n=n())
metadata$bio_reps<-bio_reps$n[match(metadata$sshc,bio_reps$sshc)]
##And area names
metadata$New_Area<-areas$New_Area[match(metadata$cluster,areas$Cluster)]
```

```{r final filtering}

# Make new phyloseq object with extended metadata
tax_mat_b<-as.matrix(taxa)
OTU_COI = otu_table(otus1, taxa_are_rows = FALSE)
TAX_S = tax_table(tax_mat_b)
p_COI = phyloseq(OTU_COI, TAX_S)

## Add metadata
sampledatas = sample_data(data.frame(metadata, row.names=metadata$root, stringsAsFactors=FALSE))
COSQ_new = merge_phyloseq(p_COI, sampledatas)
COSQ_new

## Remove cluster 2 (which was only sampled in 1 season)
COSQ_70_no_c2<-subset_samples(COSQ_new,!cluster==2)
COSQ_70_no_c2 # 6 field reps removed

#Select only samples with 3 replicates
DADAwang1<-subset_samples(COSQ_70_no_c2, bio_reps==3)
DADAwang1 # 12 field reps (6 samples) removed

### Remove OTUs from phyloseq object, that are no longer represented in any samples
DADAwang2 = filter_taxa(DADAwang1, function(x) sum(x) > 0, TRUE)
DADAwang2

# Merge field replicates
DT1<-merge_samples(DADAwang2, "sshc", fun = mean)

#Sample data needs to be remade after merging of field replicates
d<-data.frame(sample_data(DT1)[,c("cluster","season","habitat")])
#Splits rowname into habitat from third argument
d$habitat<- sapply(strsplit(as.character(rownames(d)), "_"), function(x) unlist(strsplit(x[3], "_"))[1])
#Splits rowname into season from second argument
d$season<- sapply(strsplit(as.character(rownames(d)), "_"), function(x) unlist(strsplit(x[2], "_"))[1])
d$sshc<-rownames(d)
d$sch<-paste(d$season,d$cluster,d$habitat,sep="_")
d$substrate_type<-metadata$substrate_type[match(rownames(d),metadata$sshc)]

sample_data(DT1)<-d[,c("sshc","sch","cluster","season","habitat","substrate_type")]

#Export phyloseq object as rds file
saveRDS(DT1,file="../../RDS/COI_no_c2_3reps.rds")

## Get OTU table
rep_agg<-data.frame(otu_table(DT1),check.names=F)

## Get taxonomy
tax_70<-data.frame(tax_table(DT1))
```

```{r prepare data}

# Save full abundance table for boxplots
rep_agg_abun<-rep_agg

rep_agg$Substrate<-metadata$substrate_type[match(row.names(rep_agg),metadata$sshc)]

#Subset data to sediment and water
sed<-rep_agg[rep_agg$Substrate=="sediment",]
sed<-within(sed,rm(Substrate))
wat<-rep_agg[rep_agg$Substrate=="water",]
wat<-within(wat,rm(Substrate))

## Remove empty samples
sed<-sed[rowSums(sed[])>0,] 
wat<-wat[rowSums(wat[])>0,] 

# Remove empty colums (species)
sed<-sed[colSums(sed)>0]
wat<-wat[colSums(wat)>0]

##Convert to presence/absence
sed[sed>0]<-1
wat[wat>0]<-1
```

```{r prepare boxplot}

# Subset to clusters where all three habitats were sampled
metadata$sc<-paste(metadata$season,metadata$cluster,sep="_")
rep_agg_abun$sc<-metadata$sc[match(row.names(rep_agg_abun),metadata$sshc)]
length(unique(rep_agg_abun$sc)) # 62
hab_reps<-rep_agg_abun %>% group_by(sc) %>% dplyr::summarise(n=n())
six_hreps<-filter(hab_reps,n=="6")
nrow(six_hreps) # 28
rep_6hab<-rep_agg_abun[rep_agg_abun$sc %in% six_hreps$sc, ]

##Convert to presence/absence
rep_6hab<-within(rep_6hab,rm(sc))
rep_6hab[rep_6hab>0]<-1

# Calculate MOTU richness per sample
rep_6hab$rich<-rowSums(rep_6hab)

rep_6hab$Substrate<-metadata$substrate_type[match(row.names(rep_6hab),metadata$sshc)]
rep_6hab$habitat<-metadata$habitat[match(row.names(rep_6hab),metadata$sshc)]
rep_6hab$season<-metadata$season[match(row.names(rep_6hab),metadata$sshc)]

is_outlier <- function(x) {
  return(x < quantile(x, 0.25) - 1.5 * IQR(x) | x > quantile(x, 0.75) + 1.5 * IQR(x))
}

dat <- rep_6hab %>% tibble::rownames_to_column(var="outlier") %>% group_by(habitat,season,Substrate) %>% mutate(is_outlier=ifelse(is_outlier(rich), rich, as.numeric(NA)))
dat$outlier[which(is.na(dat$is_outlier))] <- as.numeric(NA)
dat$cluster<-metadata$cluster[match(dat$outlier,metadata$sshc)]
```

```{r grouped boxplot, fig.width = 10, fig.height = 7}

ggplot(dat, aes(x=habitat, y=rich, fill=Substrate)) + 
    geom_boxplot() +
  scale_fill_manual(values = c('#d8b365','#5ab4ac'),
                     name = "", breaks = c("sediment","water"),
                     labels = c("sediment","water")) +
  xlab("") + ylab("MOTU richness")+
  facet_grid(. ~ season)+ 
  theme(legend.position = "right", text = element_text(size = 18))+
  geom_text(aes(label=cluster),na.rm=TRUE,nudge_y=0.05)
```

To test for statistically significant differences between habitats we will first look at the distribution of the data.
```{r distribution plots of raw richness, warning = FALSE}

ggplot(data = dat, aes(x = rich, group = habitat, fill=habitat)) +
    geom_density(adjust=1.5, alpha=.4) +
    scale_fill_viridis(discrete=TRUE) +
    scale_color_viridis(discrete=TRUE) +
  theme_ipsum() +
    labs(y = "", x = "Richness", title = "Distribution of Raw Richness values", fill = "Habitat")
```

Richness does look a bit left-skewed, but lets formally test it:

```{r testing skewness, include = TRUE, warning = FALSE}

skewness(dat$rich)
```

There is little skewness in the data (value was >0.5) but lets test if any habitat has a statistically different richness from the others.
Apart from the outliers we might still be able to use a parametric test (e.g. ANOVA) because 
we don't have to have normality of the single datapoints we just need homoscedascity (i.e. the residuals need to be normally distributed).
So lets try an ANOVA and check for normality of residuals.

```{r Habitat ANOVA test }

dat$habitat <- as.factor(dat$habitat)
res_aov_rich_hab <- aov(rich ~ habitat,
  data = dat)
summary(res_aov_rich_hab)
```

There is significant difference in richness between habitats!
Lets look at the residuals.

```{r Habitat ANOVA residuals, message=FALSE}

hist(res_aov_rich_hab$residuals)
qqPlot(res_aov_rich_hab$residuals, id = FALSE)
```

Plotting the residuals from the ANOVA does not clearly show that residuals are normally distributed.
We would do best to perform a non-parametric test. Let's try the Kruskal-Wallis test:

```{r Kruskal-Wallis test habitat}

kruskal.test(rich ~ habitat,
  data = dat)
```

With the non-parametric test there is still a significant difference in CO1 richness between habitats.
Lets see where the difference lies.
We will use a non-parametric post-hoc test, the Dunn test.
```{r Dunn test,message=FALSE}

dunn_res_hab <- dunnTest(rich ~ habitat,
  data = dat,
  method = "bonferroni"
)
dunn_res_hab
```

From these comparisons it looks like there is a higher richness in rocks compared to either sand or eelgrass, but after correcting for multiple testing, the differences are not significant.

Lets visualise these differences on the box plot, first across seasons:

```{r grouped boxplot w. sig diff w.o. season, fig.width = 10, fig.height = 7}

dat$Substrate <- factor(dat$Substrate)

boxplot_habitat<- ggplot(dat, aes(x=habitat, y=rich, fill=Substrate)) + 
    geom_boxplot() +
  scale_fill_manual(values = c('#d8b365','#5ab4ac'),
                     name = "", breaks = c("sediment","water"),
                     labels = c("sediment","water")) +
  xlab("") + ylab("MOTU richness")+
  theme(legend.position = "right" , text = element_text(size = 18))+
  geom_text_repel(aes(label=cluster),na.rm=TRUE,nudge_y=0.05)+
 geom_pwc(aes(group = habitat),
           tip.length = 0,
           method = "dunn_test")

ggadjust_pvalue(
  boxplot_habitat, p.adjust.method = "bonferroni",
  label = "{p.adj.signif}") #change to {p.adj.format} to get actual values
```
And then a boxplot of habitat differences split into season:

```{r grouped boxplot w. sig diff, fig.width = 12, fig.height = 8}

dat$season <- factor(dat$season, levels=c("spring", "autumn"))

boxplot_habitat_season <- ggplot(dat, aes(x=habitat, y=rich, fill=habitat)) + 
    geom_boxplot() +
  scale_fill_manual(values = c("yellowgreen", "cornflowerblue","thistle3"),
                     name = "", breaks = c("eelgrass","rocks", "sand"),
                     labels = c("eelgrass","rocks", "sand")) +
  xlab("") + ylab("MOTU richness")+
  facet_wrap(Substrate ~ season, ncol = 4)+ 
  geom_text_repel(aes(label=cluster),na.rm=TRUE,nudge_y=0.05)+
  geom_pwc(aes(group = habitat),
           tip.length = 0,
           method = "dunn_test",hide.ns = TRUE)+
  theme_classic(base_size = 18)+
  theme(legend.position = "NA" , text = element_text(size = 18))

boxplot_habitat_adj <- ggadjust_pvalue(
  boxplot_habitat_season, p.adjust.method = "bonferroni",
  label = "{p.adj.signif}") #change to {p.adj.format} to get actual values

boxplot_habitat_adj
saveRDS(boxplot_habitat_adj,"../../Plots/Boxplots/COI_habitat.rds")
```
Looks like only the eelgrass-rock comparison in autumn, sediment survives correction for multiple testing. 

Next up we are going to test if there are significant differences in ASV richness across substrates i.e. water and sediment.

```{r Substrate ANOVA test }

res_aov_rich_subst <- aov(rich ~ Substrate,
  data = dat)
summary(res_aov_rich_subst)
```

There is a significant difference in richness between sediment and water!
Lets check the residuals for normality.

```{r Substrate ANOVA residuals, message=FALSE}

hist(res_aov_rich_subst$residuals)
qqPlot(res_aov_rich_subst$residuals, id = FALSE)
```

Plotting the residuals from the anova it seems that the residuals are not normally distributed.
So we will run a Kruskal-Wallis test like before

```{r Kruskal-Wallis test habitat}

kruskal.test(rich ~ Substrate,
  data = dat)
```
There is (still) a significant difference in richness between sediment and water
Now we'll illustrate these differences on a boxplot across seasons:

```{r grouped boxplot w. substrate sig diff, fig.width = 10, fig.height = 7}

boxplot_substr <- ggplot(dat, aes(x=Substrate, y=rich, fill = habitat)) + 
    geom_boxplot() +
  scale_fill_manual(values = c("yellowgreen", "cornflowerblue","thistle3"),
                     name = "", breaks = c("eelgrass","rocks", "sand"),
                     labels = c("eelgrass","rocks", "sand")) +
  xlab("") + ylab("MOTU richness")+
  theme(legend.position = "right" , text = element_text(size = 18))+
  geom_text_repel(aes(label=cluster),na.rm=TRUE,nudge_y=0.05)+
  geom_pwc(aes(group = Substrate),
           tip.length = 0,
           method = "dunn_test")

ggadjust_pvalue(
  boxplot_substr, p.adjust.method = "bonferroni",
  label = "{p.adj.signif}") #change to {p.adj.format} to get actual values
```

Now we'll illustrate these differences on a boxplot grouped by season:

```{r grouped boxplot w. substrate sig diff w. season, fig.width = 10, fig.height = 7}

boxplot_substr_season <- ggplot(dat, aes(x=Substrate, y=rich, fill = habitat)) + 
    geom_boxplot() +
  scale_fill_manual(values = c("yellowgreen", "cornflowerblue","thistle3"),
                     name = "", breaks = c("eelgrass","rocks", "sand"),
                     labels = c("eelgrass","rocks", "sand")) +
  xlab("") + ylab("MOTU richness")+
  facet_grid(. ~ season)+ 
  theme(legend.position = "NA" , text = element_text(size = 18))+
  geom_text_repel(aes(label=cluster),na.rm=TRUE,nudge_y=0.05)+
  geom_pwc(aes(group = Substrate),
           tip.length = 0,
           method = "dunn_test")

boxplot_substr_adj <- ggadjust_pvalue(
  boxplot_substr_season, p.adjust.method = "bonferroni",
  label = "{p.adj.signif}") #change to {p.adj.format} to get actual values

boxplot_substr_adj
saveRDS(boxplot_substr_adj,"../../Plots/Boxplots/COI_substrate.rds")
```
Looks like the difference between richness in water and sediment is greater in the spring compared to the autumn. Lets test if there is a significant interaction effect between substrate and season.

```{r Substrate season interaction ANOVA test }

res_aov_rich_subst_inter <- aov(rich ~ Substrate*season,
  data = dat)
summary(res_aov_rich_subst_inter)
```
From the ANOVA results it looks like there is indeed an interaction effect of season and substrate as well as a main effect of both substrate and season.

```{r Venn function}

# Function for plotting euler diagrams with ggplot2 (adapted from: https://gist.github.com/danlooo/d23d8bcf8856c7dd8e86266097404ded)

ggeulerr <- function(combinations, show_quantities = TRUE, show_labels = FALSE, ...) {
  data <-
    eulerr::euler(combinations = combinations) %>%
    plot(quantities = show_quantities) %>%
    pluck("data")
  
  tibble() %>%
    ggplot() +
    ggforce::geom_ellipse(
      data = data$ellipses %>% as_tibble(rownames = "Set"),
      mapping = aes(x0 = h, y0 = k, a = a, b = b, angle = 0, fill = Set),
      alpha = 0.5
    ) +
    geom_text(
      data = {
        data$centers %>%
          mutate(
            label = labels %>% map2(quantities, ~ {
              if (!is.na(.x) && !is.na(.y) && show_labels) {
                paste0(.x, "\n", sprintf(.y, fmt = "%.3g"))
              } else if (!is.na(.x) && show_labels) {
                .x
              } else if (!is.na(.y)) {
                .y
              } else {
                ""
              }
            })
          )
      },
      mapping = aes(x = x, y = y, label = label), fontface="bold", size=2
    ) +
    theme(panel.grid = element_blank()) +
    coord_fixed()
}
```

```{r venn both}

# Subset to clusters where all three habitats were sampled
rep_agg$sc<-metadata$sc[match(row.names(rep_agg),metadata$sshc)]
length(unique(rep_agg$sc)) # 62
hab_reps<-rep_agg %>% group_by(sc) %>% dplyr::summarise(n=n())
three_hreps<-filter(hab_reps,n=="6")
nrow(three_hreps) # 28
both_3hab<-rep_agg[rep_agg$sc %in% three_hreps$sc, ]

# Subset to each habitat
both_3hab$habitat<-metadata$habitat[match(row.names(both_3hab),metadata$sshc)]

eel<-both_3hab[both_3hab$habitat=="eelgrass",]
rock<-both_3hab[both_3hab$habitat=="rocks",]
sand<-both_3hab[both_3hab$habitat=="sand",]

# Remove unnecessary columns
eel<-within(eel,rm(habitat,sc,Substrate))
rock<-within(rock,rm(habitat,sc,Substrate))
sand<-within(sand,rm(habitat,sc,Substrate))

##Convert to presence/absence
eel[eel>0]<-1
rock[rock>0]<-1
sand[sand>0]<-1

# Make species list for each habitat
eel<-eel[,colSums(eel)>0]
eel_sp<-colnames(eel)

rock<-rock[,colSums(rock)>0]
rock_sp<-colnames(rock)

sand<-sand[,colSums(sand)>0]
sand_sp<-colnames(sand)

both_list <- list(Eelgrass = eel_sp,Rock = rock_sp,Sand = sand_sp)

venn_both <- ggeulerr(both_list[1:3]) +
  theme_bw() + # For white background
  theme(axis.title = element_blank(), axis.text = element_blank(), axis.ticks = element_blank()) +  # To remove axis text, titles and ticks
  theme(panel.grid = element_blank()) + # To remove gridlines
  theme(panel.border = element_blank()) + # To remove plot border
  scale_fill_manual(values = c(Eelgrass = "yellowgreen", Sand = "thistle3", Rock = "cornflowerblue"))  #Setting colors for categories
  #+ guides(fill="none") # To remove legend for categories (these are shown with labels anyways)

venn_both

saveRDS(venn_both,"../../Plots/Venn/COI_70_both.rds")

rep_agg<-within(rep_agg,rm(sc))
```

```{r venn sed}

# Subset to clusters where all three habitats were sampled
metadata$sc<-paste(metadata$season,metadata$cluster,sep="_")
sed$sc<-metadata$sc[match(row.names(sed),metadata$sshc)]
length(unique(sed$sc)) # 62
hab_reps<-sed %>% group_by(sc) %>% dplyr::summarise(n=n())
three_hreps<-filter(hab_reps,n=="3")
nrow(three_hreps) # 32
sed_3hab<-sed[sed$sc %in% three_hreps$sc, ]

# Subset to each habitat
sed_3hab$habitat<-metadata$habitat[match(row.names(sed_3hab),metadata$sshc)]

eel<-sed_3hab[sed_3hab$habitat=="eelgrass",]
rock<-sed_3hab[sed_3hab$habitat=="rocks",]
sand<-sed_3hab[sed_3hab$habitat=="sand",]

eel<-within(eel,rm(habitat,sc))
rock<-within(rock,rm(habitat,sc))
sand<-within(sand,rm(habitat,sc))

# Make species list for each habitat
eel<-eel[,colSums(eel)>0]
eel_sp<-colnames(eel)

rock<-rock[,colSums(rock)>0]
rock_sp<-colnames(rock)

sand<-sand[,colSums(sand)>0]
sand_sp<-colnames(sand)

sed_list <- list(Eelgrass = eel_sp,Rock = rock_sp,Sand = sand_sp)

venn_sed <- ggeulerr(sed_list[1:3]) +
  theme_bw(base_size = 14) + # For white background
  theme(axis.title = element_blank(), axis.text = element_blank(), axis.ticks = element_blank()) +  # To remove axis text, titles and ticks
  theme(panel.grid = element_blank()) + # To remove gridlines
  theme(panel.border = element_blank()) + # To remove plot border
  scale_fill_manual(values = c(Eelgrass = "yellowgreen", Sand = "thistle3", Rock = "cornflowerblue"))  #Setting colors for categories
  #+ guides(fill="none") # To remove legend for categories (these are shown with labels anyways)

venn_sed

ggsave(filename='../../Plots/Venn/COI_70_sed.png',venn_sed)

saveRDS(venn_sed,'../../Plots/Venn/COI_70_sed.rds')

sed<-within(sed,rm(sc))
```

```{r venn wat}

# Subset to clusters where all three habitats were sampled
wat$sc<-metadata$sc[match(row.names(wat),metadata$sshc)]
length(unique(wat$sc)) # 62
hab_reps<-wat %>% group_by(sc) %>% dplyr::summarise(n=n())
three_hreps<-filter(hab_reps,n=="3")
nrow(three_hreps) # 32
wat_3hab<-wat[wat$sc %in% three_hreps$sc, ]

# Subset to each habitat
wat_3hab$habitat<-metadata$habitat[match(row.names(wat_3hab),metadata$sshc)]

eel<-wat_3hab[wat_3hab$habitat=="eelgrass",]
rock<-wat_3hab[wat_3hab$habitat=="rocks",]
sand<-wat_3hab[wat_3hab$habitat=="sand",]

eel<-within(eel,rm(habitat,sc))
rock<-within(rock,rm(habitat,sc))
sand<-within(sand,rm(habitat,sc))

# Make species list for each habitat
eel<-eel[,colSums(eel)>0]
eel_sp<-colnames(eel)

rock<-rock[,colSums(rock)>0]
rock_sp<-colnames(rock)

sand<-sand[,colSums(sand)>0]
sand_sp<-colnames(sand)

wat_list <- list(Eelgrass = eel_sp,Rock = rock_sp,Sand = sand_sp)

venn_wat <- ggeulerr(wat_list[1:3]) +
  theme_bw(base_size = 14) + # For white background
  theme(axis.title = element_blank(), axis.text = element_blank(), axis.ticks = element_blank()) +  # To remove axis text, titles and ticks
  theme(panel.grid = element_blank()) + # To remove gridlines
  theme(panel.border = element_blank()) + # To remove plot border
  scale_fill_manual(values = c(Eelgrass = "yellowgreen", Sand = "thistle3", Rock = "cornflowerblue"))  #Setting colors for categories
  #+ guides(fill="none") # To remove legend for categories (these are shown with labels anyways)

venn_wat

ggsave(filename='../../Plots/Venn/COI_70_wat.png',venn_wat)

saveRDS(venn_wat,'../../Plots/Venn/COI_70_wat.rds')

wat<-within(wat,rm(sc))
```

```{r nmds}

##Convert to presence/absence
both<-within(rep_agg,rm(Substrate))
both[both>0]<-1

#NB., the number of points n should be n > 2k + 1, and preferably higher in global non-metric MDS, and still higher in local NMDS. A k higher than 5 makes it difficult to interpret results

## Parameter settings inspired by https://rpubs.com/CPEL/NMDS

  set.seed(123)
  allmds <- metaMDS(both, distance = "jaccard", autotransform = FALSE, maxit=999, k=3, trymax=250)

  site_scrs <- as.data.frame(scores(allmds, display = "sites"))
  site_scrs$Substrate <- metadata$substrate_type[match(row.names(site_scrs), metadata$sshc)]

  cent <- aggregate(cbind(NMDS1, NMDS2) ~ Substrate, data=site_scrs, FUN = mean)
  segs <- merge(site_scrs, setNames(cent, c("Substrate", "oNMDS1", "oNMDS2")), by = "Substrate", sort=FALSE)

  cols <- c("sediment" = "#d8b365", "water" = "#5ab4ac")
          
  nmds_both<-ggplot(site_scrs, aes(x=NMDS1, y=NMDS2, colour = Substrate)) + 
  scale_color_manual(values = cols,
                     name = "", breaks = unique(metadata$substrate_type),
                     labels = unique(metadata$substrate_type)) +
  geom_point(alpha=1,size=3) +
  coord_fixed() +
  theme_classic() + 
  theme(legend.position = "right", legend.text = element_text(size = 10), legend.title = element_text(size = 8), axis.text = element_text(size = 8), axis.title.x = element_blank(),axis.title.y = element_text(size = 8),plot.margin = unit(c(0,0.4,0,0.4), 'lines'))+
  ggtitle("")+
  #set breaks on x-axis
scale_x_continuous(breaks = c(-1.0,0,1.0))+
  #set breaks on y-axis
scale_y_continuous(breaks = c(-1.5,-1,-0.5,0,0.5,1,1.5,2))
  
saveRDS(nmds_both,"../../Plots/NMDS/NMDS_COI_both.rds")
```

```{r divide by supergroup}

# Load table with supergroups
sgroups <- read.table("../../Tekstfiler/Across_barcodes/Supergroups_and_cellularity.tsv", sep='\t', header=T, comment="")

# Add supergroups
tax_70$supergroup <- sgroups$supergroup[match(tax_70$new_phylum,sgroups$new_phylum)]

## Add supergroup to Bacillariophyta_NA
tax_70$supergroup[tax_70$new_phylum=="Bacillariophyta_NA"]<-"SAR_Stramenopiles"

## Subset taxonomy table to each of three dominant kingdoms
SAR<-tax_70[which(tax_70$supergroup=="SAR_Stramenopiles"),]
plant<-tax_70[which(tax_70$supergroup=="Archaeplastids"),]
opist<-tax_70[which(tax_70$supergroup=="Opisthokonts"),]
amoeb<-tax_70[which(tax_70$supergroup=="Amoebozoans"),]

## Subset site-species tables to each of these kingdoms
sed.SAR<-sed[,names(sed) %in% row.names(SAR)]
sed.SAR<-sed.SAR[rowSums(sed.SAR[])>0,] # Remove empty samples
sed.plant<-sed[,names(sed) %in% row.names(plant)]
sed.plant<-sed.plant[rowSums(sed.plant[])>0,] # Remove empty samples
sed.opist<-sed[,names(sed) %in% row.names(opist)]
sed.opist<-sed.opist[rowSums(sed.opist[])>0,] # Remove empty samples
sed.amoeb<-sed[,names(sed) %in% row.names(amoeb)]
sed.amoeb<-sed.amoeb[rowSums(sed.amoeb[])>0,] # Remove empty samples

wat.SAR<-wat[,names(wat) %in% row.names(SAR)]
wat.SAR<-wat.SAR[rowSums(wat.SAR[])>0,] # Remove empty samples
wat.plant<-wat[,names(wat) %in% row.names(plant)]
wat.plant<-wat.plant[rowSums(wat.plant[])>0,] # Remove empty samples
wat.opist<-wat[,names(wat) %in% row.names(opist)]
wat.metaz<-wat.opist[rowSums(wat.opist[])>0,] # Remove empty samples
wat.amoeb<-wat[,names(wat) %in% row.names(amoeb)]
wat.amoeb<-wat.amoeb[rowSums(wat.amoeb[])>0,] # Remove empty samples
```

```{r nmds sed habitat}

#NB., the number of points n should be n > 2k + 1, and preferably higher in global non-metric MDS, and still higher in local NMDS. A k higher than 5 makes it difficult to interpret results

## Parameter settings inspired by https://rpubs.com/CPEL/NMDS

  cols <- c("sand"="thistle3","eelgrass"="yellowgreen","rocks"="cornflowerblue")

  set.seed(123)
  sed_nmds <- metaMDS(sed, distance = "jaccard", autotransform = FALSE, maxit=999, k=3, trymax=250)

  site_scrs <- as.data.frame(scores(sed_nmds, display = "sites"))
  site_scrs$habitat <- metadata$habitat[match(row.names(site_scrs), metadata$sshc)]
  site_scrs$season <- metadata$season[match(row.names(site_scrs), metadata$sshc)]
  
write.table(site_scrs,file="../../Tekstfiler/COI/COI_70/Jaccard_sed.txt")

  cent <- aggregate(cbind(NMDS1, NMDS2) ~ habitat, data=site_scrs, FUN = mean)
  segs <- merge(site_scrs, setNames(cent, c("habitat", "oNMDS1", "oNMDS2")), by = "habitat", sort=FALSE)

  plots_hab<-list()
  
  nmds_sed<-ggplot(site_scrs, aes(x=NMDS1, y=NMDS2, colour = habitat, shape = season)) + 
  scale_color_manual(values = cols,
                     name = "", breaks = unique(metadata$habitat),
                     labels = unique(metadata$habitat)) +
  geom_point(alpha=1,size=2) +
  coord_fixed() +
  theme_classic() + 
  theme(legend.position = "right", legend.text = element_text(size = 10), legend.title = element_blank(), axis.text = element_text(size = 8),axis.title.x = element_blank(), axis.title.y = element_blank(),plot.margin = unit(c(0,0.4,0,0.4), 'lines'))+
  ggtitle("")+
  #set breaks on x-axis
scale_x_continuous(breaks = c(-1.0,0,1.0))+
  #set breaks on y-axis
scale_y_continuous(breaks = c(-1.5,-1,-0.5,0,0.5,1,1.5,2))

saveRDS(nmds_sed,"../../Plots/NMDS/NMDS_70_hab_sed.rds")
```

##PERMANOVA analysis
```{r permanova sed}

dist<-vegdist(sed,method="jaccard")

#Make new metadata table for sediment
meta<-as.data.frame(row.names(sed))
row.names(meta)<-meta$`row.names(sed)`
meta$habitat<-metadata$habitat[match(row.names(meta),metadata$sshc)]
meta<-within(meta,rm(`row.names(sed)`))

#Test for homogeneity of multivariate dispersions
b<-betadisper(dist,meta$habitat)
permutest(b)

# There is not homogeneity of dispersions

#Perform a PERMANOVA test
permanova<-adonis2(formula = dist ~ habitat, data = meta)
permanova

# There is a significant effect of habitat (R2=0.05, p=0.001)

## Same test for season

#Test for homogeneity of multivariate dispersions
meta$season<-metadata$season[match(row.names(meta),metadata$sshc)]
b<-betadisper(dist,meta$season)
permutest(b)

# There is not homogeneity of dispersions (p=0.04)

#Perform a PERMANOVA test
permanova<-adonis2(formula = dist ~ season, data = meta)
permanova

# There is a significant effect of season (R2=0.02, p=0.001)
```

```{r nmds wat habitat}

#NB., the number of points n should be n > 2k + 1, and preferably higher in global non-metric MDS, and still higher in local NMDS. A k higher than 5 makes it difficult to interpret results

## Parameter settings inspired by https://rpubs.com/CPEL/NMDS

  set.seed(123)
  wat_nmds <- metaMDS(wat, distance = "jaccard", autotransform = FALSE, maxit=999, k=3, trymax=250)

  site_scrs_w <- as.data.frame(scores(wat_nmds, display = "sites"))
  site_scrs_w$habitat <- metadata$habitat[match(row.names(site_scrs_w), metadata$sshc)]
  site_scrs_w$season <- metadata$season[match(row.names(site_scrs_w), metadata$sshc)]
  
  write.table(site_scrs_w,file="../../Tekstfiler/COI/COI_70/Jaccard_wat.txt")

  cent <- aggregate(cbind(NMDS1, NMDS2) ~ habitat, data=site_scrs_w, FUN = mean)
  segs <- merge(site_scrs_w, setNames(cent, c("habitat", "oNMDS1", "oNMDS2")), by = "habitat", sort=FALSE)

  plots_hab_w<-list()
  
  nmds_wat<-ggplot(site_scrs_w, aes(x=NMDS1, y=NMDS2, colour = habitat, shape = season)) + 
  scale_color_manual(values = cols,
                     name = "", breaks = unique(metadata$habitat),
                     labels = unique(metadata$habitat)) +
  geom_point(alpha=1,size=2) +
  coord_fixed() +
  theme_classic() + 
  theme(legend.position = "right", legend.text = element_text(size = 10), legend.title = element_blank(), axis.text = element_text(size = 8),axis.title.x = element_blank(), axis.title.y = element_blank(),plot.margin = unit(c(0,0.4,0,0.4), 'lines'))+
  ggtitle("")+
  #set breaks on x-axis
scale_x_continuous(breaks = c(-1.0,0,1.0))+
  #set breaks on y-axis
scale_y_continuous(breaks = c(-1.5,-1,-0.5,0,0.5,1,1.5,2))

saveRDS(nmds_wat,"../../Plots/NMDS/NMDS_70_hab_wat.rds")
```

##PERMANOVA analysis
```{r permanova wat}

dist<-vegdist(wat,method="jaccard")

#Make new metadata table for water
meta<-as.data.frame(row.names(wat))
row.names(meta)<-meta$`row.names(wat)`
meta$habitat<-metadata$habitat[match(row.names(meta),metadata$sshc)]
meta<-within(meta,rm(`row.names(wat)`))

#Test for homogeneity of multivariate dispersions
b<-betadisper(dist,meta$habitat)
permutest(b)

# There is not homogeneity of dispersions

#Perform a PERMANOVA test
permanova<-adonis2(formula = dist ~ habitat, data = meta)
permanova

# There is a significant effect of habitat (R2=0.03, p=0.001)

## Same test for season

#Test for homogeneity of multivariate dispersions
meta$season<-metadata$season[match(row.names(meta),metadata$sshc)]
b<-betadisper(dist,meta$season)
permutest(b)

# There is not homogeneity of dispersions

#Perform a PERMANOVA test
permanova<-adonis2(formula = dist ~ season, data = meta)
permanova

# There is a significant effect of season (R2=0.07, p=0.001)
```

```{r nmds sed salinity}

## Get salinities
coord<-read.table("../../Tekstfiler/Across_barcodes/merged_metadata_230427.txt", sep="\t", row.names=1, header=T, check.names=FALSE)
metadata$new_name<-paste(metadata$season,metadata$cluster,metadata$habitat,sep="_")
metadata$salinity <- coord$Salinity[match(metadata$new_name,row.names(coord))]

site_scrs$salinity <- metadata$salinity[match(row.names(site_scrs), metadata$sshc)]

cent <- aggregate(cbind(NMDS1, NMDS2) ~ salinity, data=site_scrs, FUN = mean)
segs <- merge(site_scrs, setNames(cent, c("salinity", "oNMDS1", "oNMDS2")), by = "salinity", sort=FALSE)

  # set breaks and colours
breaks <- c(7,11,17,22,27,33)
breaks

cols <- RColorBrewer::brewer.pal(9, "RdYlBu")

plots<-list()

  nmds_sal<-ggplot(site_scrs, aes(x=NMDS1, y=NMDS2)) + 
  scale_colour_stepsn(colours = rev(cols),
                    breaks = breaks,
                    name = "Salinity") +
  geom_point(alpha=1, size=2, aes(colour = salinity, shape = season)) +
  coord_fixed() +
  theme_classic() + 
  theme(legend.position = "right", legend.text = element_text(size = 10), legend.title = element_text(size = 12), axis.text = element_text(size = 8),axis.title.x = element_blank(), axis.title.y = element_blank(),plot.margin = unit(c(0,0.4,0,0.4), 'lines'))+
  ggtitle("") +
  #set breaks on x-axis
scale_x_continuous(breaks = c(-1.0,0,1.0))+
  #set breaks on y-axis
scale_y_continuous(breaks = c(-1.5,-1,-0.5,0,0.5,1,1.5,2))
  
saveRDS(nmds_sal,"../../Plots/NMDS/NMDS_70_sal_sed.rds")
```

```{r nmds wat salinity}

## Get salinities
site_scrs_w$salinity <- metadata$salinity[match(row.names(site_scrs_w), metadata$sshc)]

cent <- aggregate(cbind(NMDS1, NMDS2) ~ salinity, data=site_scrs_w, FUN = mean)
segs <- merge(site_scrs_w, setNames(cent, c("salinity", "oNMDS1", "oNMDS2")), by = "salinity", sort=FALSE)

  # set breaks and colours
breaks <- c(7,11,17,22,27,33)
breaks

cols <- RColorBrewer::brewer.pal(9, "RdYlBu")

plots_wat<-list()

  nmds_sal_w<-ggplot(site_scrs_w, aes(x=NMDS1, y=NMDS2)) + 
  scale_colour_stepsn(colours = rev(cols),
                    breaks = breaks,
                    name = "Salinity") +
  geom_point(alpha=1, size=2, aes(colour = salinity, shape = season)) +
  coord_fixed() +
  theme_classic() + 
  theme(legend.position = "right", legend.text = element_text(size = 10), legend.title = element_text(size = 12), axis.text = element_text(size = 8),axis.title.x = element_blank(), axis.title.y = element_blank(),plot.margin = unit(c(0,0.4,0,0.4), 'lines'))+
  ggtitle("") +
  #set breaks on x-axis
scale_x_continuous(breaks = c(-1.0,0,1.0))+
  #set breaks on y-axis
scale_y_continuous(breaks = c(-1.5,-1,-0.5,0,0.5,1,1.5,2))
  
saveRDS(nmds_sal_w,"../../Plots/NMDS/NMDS_70_sal_wat.rds")
```

```{r nmds sed salinity supergroup}

datasets<-list(sed.SAR,sed.opist,sed.amoeb,sed.plant)
names<-c("SAR_Stramenopiles","Opisthokonts","Amoebozoans","Archaeplastids")

#NB., the number of points n should be n > 2k + 1n>2k+1, and preferably higher in global non-metric MDS, and still higher in local NMDS. A k higher than 5 makes it difficult to interpret results

## Parameter settings inspired by https://rpubs.com/CPEL/NMDS

plots<-list()

for (i in 1:length(datasets)){

  set.seed(123)
  sed_nmds <- metaMDS(datasets[[i]], distance = "jaccard", autotransform = FALSE, maxit=999, k=4, trymax=250)

  site_scrs <- as.data.frame(scores(sed_nmds, display = "sites"))
  site_scrs$season <- metadata$season[match(row.names(site_scrs), metadata$sshc)]
  
  ## Get salinities
  site_scrs$salinity <- metadata$salinity[match(row.names(site_scrs), metadata$sshc)]

  cent <- aggregate(cbind(NMDS1, NMDS2) ~ salinity, data=site_scrs, FUN = mean)
  segs <- merge(site_scrs, setNames(cent, c("salinity", "oNMDS1", "oNMDS2")), by = "salinity", sort=FALSE)

  # set breaks and colours
  breaks <- c(7,11,17,22,27,33)
  breaks

  cols <- RColorBrewer::brewer.pal(9, "RdYlBu")

  nmds_sal<-ggplot(site_scrs, aes(x=NMDS1, y=NMDS2)) + 
  scale_colour_stepsn(colours = rev(cols),
                    breaks = breaks,
                    name = "Salinity") +
  geom_point(alpha=1, aes(colour = salinity, shape = season)) +
  coord_fixed() +
  theme_classic() + 
  theme(legend.position = "right", legend.text = element_text(size = 8), legend.title = element_text(size = 8), axis.text = element_text(size = 8),axis.title.x = element_blank(),axis.title.y = element_text(size = 8), plot.title = element_text(size = 10), plot.margin = unit(c(0,0.4,0,0.4), 'lines'))+
  ggtitle(names[[i]])
  
  plots[[i]]<-nmds_sal
}  

nmds_sal
```

```{r nmds wat salinity kin}

datasets<-list(wat.SAR,wat.opist,wat.amoeb,wat.plant)

#NB., the number of points n should be n > 2k + 1n>2k+1, and preferably higher in global non-metric MDS, and still higher in local NMDS. A k higher than 5 makes it difficult to interpret results

## Parameter settings inspired by https://rpubs.com/CPEL/NMDS

plots_wat<-list()

for (i in 1:length(datasets)){

  set.seed(123)
  wat_nmds <- metaMDS(datasets[[i]], distance = "jaccard", autotransform = FALSE, maxit=999, k=4, trymax=250)

  site_scrs <- as.data.frame(scores(wat_nmds, display = "sites"))
  site_scrs$season <- metadata$season[match(row.names(site_scrs), metadata$sshc)]
  
  ## Get salinities
  site_scrs$salinity <- metadata$salinity[match(row.names(site_scrs), metadata$sshc)]

  cent <- aggregate(cbind(NMDS1, NMDS2) ~ salinity, data=site_scrs, FUN = mean)
  segs <- merge(site_scrs, setNames(cent, c("salinity", "oNMDS1", "oNMDS2")), by = "salinity", sort=FALSE)

  # set breaks and colours
  breaks <- c(7,11,17,22,27,33)
  breaks

  cols <- RColorBrewer::brewer.pal(9, "RdYlBu")

  nmds_sal<-ggplot(site_scrs, aes(x=NMDS1, y=NMDS2)) + 
  scale_colour_stepsn(colours = rev(cols),
                    breaks = breaks,
                    name = "Salinity") +
  geom_point(alpha=1, aes(colour = salinity, shape = season)) +
  coord_fixed() +
  theme_classic() + 
  theme(legend.position = "right", legend.text = element_text(size = 8), legend.title = element_text(size = 8), axis.text = element_text(size = 6),axis.title.x = element_blank(),axis.title.y = element_text(size = 8), plot.title = element_text(size = 10),plot.margin = unit(c(0,0.4,0,0.4), 'lines'))+
  ggtitle(names[[i]])
  
  plots_wat[[i]]<-nmds_sal
}  

ggsave(file="../../Plots/NMDS/NMDS_70_both_sal_supergroups.png",ggarrange(plots[[1]],plots_wat[[1]],plots[[2]],plots_wat[[2]],plots[[3]],plots_wat[[3]],plots[[4]],plots_wat[[4]],ncol=2,nrow=4,align="h",common.legend=TRUE,legend="right"),width = 125,
       height = 150,
       units = "mm",
       dpi = 200)
```