---
title: "MOTU analyses"
author: "Eva Egelyng Sigsgaard"
date: "2025-02-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, message=FALSE}
library(dplyr)
library(vegan)
library(RColorBrewer)
library(ggplot2)
library(ggpubr)
library(ggrepel)
library(phyloseq)
library(hrbrthemes)
library(viridis)
library(e1071)
library(FSA)
library(car)
library(ggVennDiagram)
library(ggvenn)
library(eulerr)
library(tidyverse)
library(ggforce)
```

```{r load data}

# Load rarefied dataset (MOTU dataset)
DT1<-readRDS("../../RDS/16S_no_c2_3reps.rds")

## Get OTU table
rep_agg<-data.frame(otu_table(DT1),check.names=F)

## Get sample data
d<-data.frame(sample_data(DT1),check.names=F)
```

```{r combine substrates}

# Add column indicating season, sampling site and habitat type
rep_agg$sch<-d$sch[match(row.names(rep_agg),row.names(d))]

# Aggregate sediment and water samples from same sampling event
rep_agg_comb<-aggregate(. ~ sch, rep_agg, sum)
row.names(rep_agg_comb)<-rep_agg_comb$sch
rep_agg_comb<-within(rep_agg_comb,rm(sch))

# Save abundance table for boxplots
rep_agg_abun<-rep_agg_comb

##Convert to presence/absence
rep_agg_comb[rep_agg_comb>0]<-1
```

```{r prepare boxplots}

# Subset to clusters where all three habitats were sampled
d$sc<-paste(d$season,d$cluster,sep="_")
rep_agg_abun$sc<-d$sc[match(row.names(rep_agg_abun),d$sch)]
length(unique(rep_agg_abun$sc)) # 62
hab_reps<-rep_agg_abun %>% group_by(sc) %>% summarise(n=n())
three_hreps<-filter(hab_reps,n=="3")
nrow(three_hreps) # 32
rep_3hab<-rep_agg_abun[rep_agg_abun$sc %in% three_hreps$sc, ]

##Convert to presence/absence
rep_3hab<-within(rep_3hab,rm(sc))
rep_3hab[rep_3hab>0]<-1

# Calculate MOTU richness per sample
rep_3hab$rich<-rowSums(rep_3hab)

rep_3hab$habitat<-d$habitat[match(row.names(rep_3hab),d$sch)]
rep_3hab$season<-d$season[match(row.names(rep_3hab),d$sch)]

is_outlier <- function(x) {
  return(x < quantile(x, 0.25) - 1.5 * IQR(x) | x > quantile(x, 0.75) + 1.5 * IQR(x))
}

dat <- rep_3hab %>% tibble::rownames_to_column(var="outlier") %>% group_by(habitat,season) %>% mutate(is_outlier=ifelse(is_outlier(rich), rich, as.numeric(NA)))
dat$outlier[which(is.na(dat$is_outlier))] <- as.numeric(NA)
dat$cluster<-d$cluster[match(dat$outlier,d$sch)]
```

```{r grouped boxplot, fig.width = 10, fig.height = 7}

ggplot(dat, aes(x=habitat, y=rich)) + 
    geom_boxplot() +
  xlab("") + ylab("MOTU richness")+
  facet_grid(. ~ season)+ 
  theme(legend.position = "right", text = element_text(size = 18))+
  geom_text(aes(label=cluster),na.rm=TRUE,nudge_y=0.05)
```
To test for statistically significant differences between habitats. we will first look at the distribution of the data.
```{r distribution plots of raw richness, warning = FALSE}

ggplot(data = dat, aes(x = rich, group = habitat, fill=habitat)) +
    geom_density(adjust=1.5, alpha=.4) +
    scale_fill_viridis(discrete=TRUE) +
    scale_color_viridis(discrete=TRUE) +
  theme_ipsum() +
    labs(y = "", x = "Richness", title = "Distribution of Raw Richness values", fill = "Habitat")
```

Richness does not look skewed, but lets formally test it:

```{r testing skewness, include = TRUE, warning = FALSE}

skewness(dat$rich)
```

There is a little skewness in the data (value was >0.5), so lets test if any habitat has a statistically different richness from the others.
Apart from the outliers we might still be able to use a parametric test (e.g. ANOVA) because 
we don't have to have normality of the single datapoints we just need homoscedascity (i.e. the residuals need to be normally distributed).
So lets try an ANOVA and check for normality of residuals.

```{r Habitat ANOVA test }

dat$habitat <- as.factor(dat$habitat)
res_aov_rich_hab <- aov(rich ~ habitat,
  data = dat)
summary(res_aov_rich_hab)
```

There is significant difference in richness between habitats!
Lets look at the residuals.

```{r Habitat ANOVA residuals, message=FALSE}

hist(res_aov_rich_hab$residuals)
qqPlot(res_aov_rich_hab$residuals, id = FALSE)
```

Plotting the residuals from the ANOVA does not clearly show that residuals are normally distributed.
We would do best to perform a non-parametric test. Let's try the Kruskal-Wallis test:

```{r Kruskal-Wallis test habitat}

kruskal.test(rich ~ habitat,
  data = dat)
```

With the non-parametric test there is a significant difference in richness between habitats.
But lets see where the difference lies.
We will use a non-parametric post-hoc test, the Dunn test.
```{r Dunn test,message=FALSE}

dunn_res_hab <- dunnTest(rich ~ habitat,
  data = dat,
  method = "bonferroni"
)
dunn_res_hab
```

From these comparisons it looks like there is highest richness in rocks, second-highest in eelgrass and lowest in sand, even after correcting for multiple testing.

Lets visualise these differences on the box plot, first across seasons:

```{r grouped boxplot w. sig diff w.o. season, fig.width = 10, fig.height = 7}

boxplot_habitat<- ggplot(dat, aes(x=habitat, y=rich)) + 
    geom_boxplot() +
  xlab("") + ylab("MOTU richness")+
  theme(legend.position = "right" , text = element_text(size = 18))+
  geom_text_repel(aes(label=cluster),na.rm=TRUE,nudge_y=0.05)+
 geom_pwc(aes(group = habitat),
           tip.length = 0,
           method = "dunn_test")

ggadjust_pvalue(
  boxplot_habitat, p.adjust.method = "bonferroni",
  label = "{p.adj.signif}") #change to {p.adj.format} to get actual values
```
And then a boxplot of habitat differences split into season:

```{r grouped boxplot w. sig diff, fig.width = 12, fig.height = 8}

dat$season <- factor(dat$season, levels=c("spring", "autumn"))

cols <- c("sand"="thistle3","eelgrass"="yellowgreen","rocks"="cornflowerblue")

boxplot_habitat_season <- ggplot(dat, aes(x=habitat, y=rich, fill=habitat)) + 
  geom_boxplot() +
  scale_fill_manual(values = cols,
                     name = "", breaks = unique(d$habitat),
                     labels = unique(d$habitat)) +
  xlab("") + ylab("MOTU richness")+
  facet_wrap(. ~ season, ncol = 4)+ 
  theme(legend.position = "NA" , text = element_text(size = 18))+
  geom_text_repel(aes(label=cluster),na.rm=TRUE,nudge_y=0.05)+
  geom_pwc(aes(group = habitat),
           tip.length = 0,
           method = "dunn_test")

boxplot_habitat_adj <- ggadjust_pvalue(
  boxplot_habitat_season, p.adjust.method = "bonferroni",
  label = "{p.adj.signif}") #change to {p.adj.format} to get actual values

saveRDS(boxplot_habitat_adj,"../../Plots/Boxplots/16S_habitat_comb.rds")
```
Looks like only the sand-rock comparison survives correction for multiple testing. 

```{r nmds habitat}

#NB., the number of points n should be n > 2k + 1, and preferably higher in global non-metric MDS, and still higher in local NMDS. A k higher than 5 makes it difficult to interpret results

## Parameter settings inspired by https://rpubs.com/CPEL/NMDS

  set.seed(123)
  comb_nmds <- metaMDS(rep_agg_comb, distance = "jaccard", autotransform = FALSE, maxit=999, k=3, trymax=250)

  site_scrs <- as.data.frame(scores(comb_nmds, display = "sites"))
  site_scrs$habitat <- d$habitat[match(row.names(site_scrs), d$sch)]
  site_scrs$season <- d$season[match(row.names(site_scrs), d$sch)]
  
 # write.table(site_scrs,file="../../Tekstfiler/Jaccard_16S_comb.txt")

  cent <- aggregate(cbind(NMDS1, NMDS2) ~ habitat, data=site_scrs, FUN = mean)
  segs <- merge(site_scrs, setNames(cent, c("habitat", "oNMDS1", "oNMDS2")), by = "habitat", sort=FALSE)

  plots_hab<-list()
  
  nmds_sed<-ggplot(site_scrs, aes(x=NMDS1, y=NMDS2, colour = habitat, shape = season)) + 
  scale_color_manual(values = cols,
                     name = "", breaks = unique(d$habitat),
                     labels = unique(d$habitat)) +
  geom_point(alpha=1,size=2) +
  coord_fixed() +
  theme_classic() + 
  theme(legend.position = "right", legend.text = element_text(size = 10), legend.title = element_blank(), axis.text = element_text(size = 8),axis.title.x = element_blank(), axis.title.y = element_blank(),plot.margin = unit(c(0,0.4,0,0.4), 'lines'))+
  ggtitle("")+
  #set breaks on x-axis
scale_x_continuous(breaks = c(-1.0,0,1.0))+
  #set breaks on y-axis
scale_y_continuous(breaks = c(-1.5,-1,-0.5,0,0.5,1,1.5,2))

  saveRDS(nmds_sed,"../../Plots/NMDS/NMDS_16S_hab_comb.rds")
```

##PERMANOVA analysis
```{r permanova sed}

dist<-vegdist(rep_agg_comb,method="jaccard")

#Make new metadata table for sediment
meta<-as.data.frame(row.names(rep_agg_comb))
row.names(meta)<-meta$`row.names(rep_agg_comb)`
meta$habitat<-d$habitat[match(row.names(meta),d$sch)]
meta<-within(meta,rm(`row.names(rep_agg_comb)`))

#Test for homogeneity of multivariate dispersions
b<-betadisper(dist,meta$habitat)
permutest(b)

# There is a significant difference in multivariate dispersion
  
#Perform a PERMANOVA test
permanova<-adonis2(formula = dist ~ habitat, data = meta)
permanova

# There is a significant difference between habitats

## Same test for season

#Test for homogeneity of multivariate dispersions
meta$season<-d$season[match(row.names(meta),d$sch)]
b<-betadisper(dist,meta$season)
permutest(b)

# There is no significant difference in multivariate dispersion

#Perform a PERMANOVA test
permanova<-adonis2(formula = dist ~ season, data = meta)
permanova

# There is a significant difference between seasons
```

```{r nmds sed salinity}

## Get salinities
coord<-read.table("../../Tekstfiler/Across_barcodes/merged_metadata_230427.txt", sep="\t", row.names=1, header=T, check.names=FALSE)

site_scrs$salinity <- coord$Salinity[match(row.names(site_scrs), row.names(coord))]

cent <- aggregate(cbind(NMDS1, NMDS2) ~ salinity, data=site_scrs, FUN = mean)
segs <- merge(site_scrs, setNames(cent, c("salinity", "oNMDS1", "oNMDS2")), by = "salinity", sort=FALSE)

# set breaks and colours
breaks <- c(7,11,17,22,27,33)

cols <- RColorBrewer::brewer.pal(9, "RdYlBu")

plots<-list()

nmds_sal<-ggplot(site_scrs, aes(x=NMDS1, y=NMDS2)) + 
  scale_colour_stepsn(colours = rev(cols),
                    breaks = breaks,
                    name = "Salinity") +
  geom_point(alpha=1, size=2, aes(colour = salinity, shape = season)) +
  coord_fixed() +
  theme_classic() + 
  theme(legend.position = "right", legend.text = element_text(size = 10), legend.title = element_text(size = 12), axis.text = element_text(size = 8),axis.title.x = element_blank(), axis.title.y = element_blank(),plot.margin = unit(c(0,0.4,0,0.4), 'lines'))+
  ggtitle("") +
  #set breaks on x-axis
scale_x_continuous(breaks = c(-1.0,0,1.0))+
  #set breaks on y-axis
scale_y_continuous(breaks = c(-1.5,-1,-0.5,0,0.5,1,1.5,2))
  
saveRDS(nmds_sal,"../../Plots/NMDS/NMDS_16S_sal_comb.rds")
```
