---
title: "18S analyses"
author: "Eva Egelyng Sigsgaard"
date: "2023-09-29"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,fig.path = "images/")
```

```{r libraries, message=FALSE}
library(dplyr)
library(vegan)
library(RColorBrewer)
library(ggplot2)
library(ggpubr)
library(ggrepel)
library(phyloseq)
library(tibble)
library(hrbrthemes)
library(viridis)
library(e1071)
library(car)
library(FSA)
library(ggVennDiagram)
library(ggvenn)
library(eulerr)
library(tidyverse)
library(ggforce)
```

```{r load data}

## Get rel abun OTU table
#otus1<-read.table("../Tekstfiler/18S/otu_rel_hel.txt", sep="\t", row.names = 1, header=T)

## Get rarefied OTU table
otus1<-read.table("../../Tekstfiler/18S/otu_rarefy.txt")

## Get tax table
taxa<-read.table("../../Tekstfiler/18S/tax_rarefy.txt", sep="\t", row.names = 1, header=T)

##Get metadata
metadata<-read.table("../../Tekstfiler/18S/metadata_rarefy.txt", sep="\t", header=T)

#Create extra metadata variables
metadata$sshc<-paste(metadata$substrate_type, metadata$season, metadata$habitat, metadata$cluster, sep="_")
##Including the number of field reps per sample
bio_reps<-metadata %>% group_by(sshc) %>% dplyr::summarise(n=n())
metadata$bio_reps<-bio_reps$n[match(metadata$sshc,bio_reps$sshc)]
```

```{r final filtering}

###Make phyloseq object
tax_mat_b<-as.matrix(taxa)
OTU = otu_table(otus1, taxa_are_rows = FALSE)
TAX_b = tax_table(tax_mat_b)
p_NCBI = phyloseq(OTU, TAX_b)

##Add metadata
sampledata = sample_data(data.frame(metadata, row.names=metadata$sample_root, stringsAsFactors=FALSE))
DADAwang1 = merge_phyloseq(p_NCBI, sampledata)
DADAwang1

##Remove cluster 2, which was only sampled in one season
DADAwang1<-subset_samples(DADAwang1, !cluster==2)
DADAwang1 # 6 field reps removed

#Select only samples with 3 replicates
DADAwang1<-subset_samples(DADAwang1, bio_reps==3)
DADAwang1 # 4 field reps removed

# Merge field replicates
DT1<-merge_samples(DADAwang1, "sshc", fun = mean)

#Sample data needs to be remade after merging of field replicates
d<-data.frame(sample_data(DT1)[,c("cluster","season","habitat")])
#Splits rowname into habitat from third argument
d$habitat<- sapply(strsplit(as.character(rownames(d)), "_"), function(x) unlist(strsplit(x[3], "_"))[1])
#Splits rowname into season from second argument
d$season<- sapply(strsplit(as.character(rownames(d)), "_"), function(x) unlist(strsplit(x[2], "_"))[1])
d$sshc<-rownames(d)
d$sch<-paste(d$season,d$cluster,d$habitat,sep="_")
d$substrate_type<-metadata$substrate_type[match(rownames(d),metadata$sshc)]

sample_data(DT1)<-d[,c("sshc","sch","cluster","season","habitat","substrate_type")]

#Export phyloseq object as rds file
saveRDS(DT1,file="../../RDS/18S_no_c2_3reps.rds")

## Get cleaned OTU table
rep_agg<-data.frame(otu_table(DT1),check.names=F)

## Get taxonomy
tax<-data.frame(tax_table(DT1))
```

```{r prepare data}

# Save object for boxplots
rep_agg_pa<-rep_agg
rep_agg$Substrate<-metadata$substrate_type[match(row.names(rep_agg),metadata$sshc)]

#Subset data to sediment and water
sed<-rep_agg[rep_agg$Substrate=="sediment",]
sed<-within(sed,rm(Substrate))
wat<-rep_agg[rep_agg$Substrate=="water",]
wat<-within(wat,rm(Substrate))

## Remove empty samples
sed<-sed[rowSums(sed[])>0,] 
wat<-wat[rowSums(wat[])>0,] 

# Remove empty colums (species)
sed<-sed[colSums(sed)>0]
wat<-wat[colSums(wat)>0]

##Convert to presence/absence
sed[sed>0]<-1
wat[wat>0]<-1
```

```{r prepare boxplots}

# Subset to clusters where all three habitats were sampled
metadata$sc<-paste(metadata$season,metadata$cluster,sep="_")
rep_agg_pa$sc<-metadata$sc[match(row.names(rep_agg_pa),metadata$sshc)]
length(unique(rep_agg_pa$sc)) # 62
hab_reps<-rep_agg_pa %>% group_by(sc) %>% dplyr::summarise(n=n())
six_hreps<-filter(hab_reps,n=="6")
nrow(six_hreps) # 32
rep_6hab<-rep_agg_pa[rep_agg_pa$sc %in% six_hreps$sc, ]

##Convert to presence/absence
rep_6hab<-within(rep_6hab,rm(sc))
rep_6hab[rep_6hab>0]<-1

# Calculate MOTU richness per sample
rep_6hab$rich<-rowSums(rep_6hab)

rep_6hab$Substrate<-metadata$substrate_type[match(row.names(rep_6hab),metadata$sshc)]
rep_6hab$habitat<-metadata$habitat[match(row.names(rep_6hab),metadata$sshc)]
rep_6hab$season<-metadata$season[match(row.names(rep_6hab),metadata$sshc)]

is_outlier <- function(x) {
  return(x < quantile(x, 0.25) - 1.5 * IQR(x) | x > quantile(x, 0.75) + 1.5 * IQR(x))
}

dat <- rep_6hab %>% tibble::rownames_to_column(var="outlier") %>% group_by(habitat,season,Substrate) %>% mutate(is_outlier=ifelse(is_outlier(rich), rich, as.numeric(NA)))
dat$outlier[which(is.na(dat$is_outlier))] <- as.numeric(NA)
dat$cluster<-metadata$cluster[match(dat$outlier,metadata$sshc)]
```

```{r grouped boxplot, fig.width = 10, fig.height = 7}

ggplot(dat, aes(x=habitat, y=rich, fill=Substrate)) + 
    geom_boxplot() +
  scale_fill_manual(values = c('#d8b365','#5ab4ac'),
                     name = "", breaks = c("sediment","water"),
                     labels = c("sediment","water")) +
  xlab("") + ylab("MOTU richness")+
  facet_grid(. ~ season)+ 
  theme(legend.position = "right", text = element_text(size = 18))+
  geom_text_repel(aes(label=cluster),na.rm=TRUE,nudge_y=0.05)
```
To test for statistically significant differences between habitats. we will first look at the distribution of the data.
```{r distribution plots of raw richness, warning = FALSE}

ggplot(data = dat, aes(x = rich, group = habitat, fill=habitat)) +
    geom_density(adjust=1.5, alpha=.4) +
    scale_fill_viridis(discrete=TRUE) +
    scale_color_viridis(discrete=TRUE) +
  theme_ipsum() +
    labs(y = "", x = "Richness", title = "Distribution of Raw Richness values", fill = "Habitat")
```

Richness does not seem to be skewed, but lets formally test it:

```{r testing skewness, include = TRUE, warning = FALSE}

skewness(dat$rich)
```

There is no skewness in the data (value was <0.5 and >-0.5), so lets test if any habitat has a statistically different richness from the others.

Lets try an ANOVA and check for normality of residuals.

```{r Habitat ANOVA test }

dat$habitat <- as.factor(dat$habitat)
res_aov_rich_hab <- aov(rich ~ habitat,
  data = dat)
summary(res_aov_rich_hab)
```

There is significant difference in richness between habitats!
Lets look at the residuals.

```{r Habitat ANOVA residuals, message=FALSE}

hist(res_aov_rich_hab$residuals)
qqPlot(res_aov_rich_hab$residuals, id = FALSE)
```

Plotting the residuals from the anova confirms that the residuals are normally distributed.
Now lets find out which specific habitats are different from which.
Because of multiple testing we will do that with a Tukey's Honest Significant Difference (HSD) test. As opposed to a t-test, the Tukey HSD has a "built in" correction for multiple testing.
```{r Tukey HSD test}

TukeyHSD(res_aov_rich_hab)
```

From these comparisons it looks like there is a higher richness in rocks compared to either sand or eelgrass but no significant difference in richness between sand and eelgrass. This is illustrated in the plotted output below:
```{r Tukey HSD test output plot }

plot(TukeyHSD(res_aov_rich_hab))
```

Lets visualise these differences on the boxplot, first across seasons:
*Note*: unfortunately when using facet_wrap() as is done in the next plot with season included as a grouping variable, you cannot use a built-in Tukey HSD test and account for all comparisons (it only accounts for comparisons in one plot of the facets) for that reason I have chosen a t-test with a bonferroni correction. It is a little conservative but with a small number of comparisons it is argued as a preferred option and makes almost no difference from Tukey HSD.

```{r grouped boxplot w. sig diff w.o. season, fig.width = 10, fig.height = 7}

dat$Substrate <- factor(dat$Substrate)

boxplot_habitat <- ggplot(dat, aes(x=habitat, y=rich, fill=Substrate)) + 
    geom_boxplot() +
  scale_fill_manual(values = c('#d8b365','#5ab4ac'),
                     name = "", breaks = c("sediment","water"),
                     labels = c("sediment","water")) +
  xlab("") + ylab("MOTU richness")+
  theme(legend.position = "right" , text = element_text(size = 18))+
  geom_text_repel(aes(label=cluster),na.rm=TRUE,nudge_y=0.05)+
   geom_pwc(aes(group = habitat),
           tip.length = 0,
           method = "t_test")

ggadjust_pvalue(
  boxplot_habitat, p.adjust.method = "bonferroni",
  label = "{p.adj.signif}") #change to {p.adj.format} to get actual values
```
And then a boxplot of habitat differences split into season 

```{r grouped boxplot w. sig diff, fig.width = 12, fig.height = 8}

dat$season <- factor(dat$season, levels=c("spring", "autumn"))

boxplot_habitat_season <- ggplot(dat, aes(x=habitat, y=rich, fill=habitat)) + 
  geom_boxplot() +
  scale_fill_manual(values = c("yellowgreen", "cornflowerblue","thistle3"),
                     name = "", breaks = c("eelgrass","rocks", "sand"),
                     labels = c("eelgrass","rocks", "sand")) +
  xlab("") + ylab("MOTU richness")+
  geom_text_repel(aes(label=cluster),na.rm=TRUE,nudge_y=0.05)+
  facet_wrap(Substrate ~ season, ncol = 4)+
    geom_pwc(aes(group = habitat),
           tip.length = 0,
           method = "t_test",hide.ns = TRUE)+
  theme_classic(base_size = 18)+
  theme(legend.position = "top" , text = element_text(size = 18))

boxplot_habitat_adj <- ggadjust_pvalue(
  boxplot_habitat_season, p.adjust.method = "bonferroni",
  label = "{p.adj.signif}") #change to {p.adj.format} to get actual values

saveRDS(boxplot_habitat_adj,"../../Plots/Boxplots/18S_habitat.rds")
```
From the plot we can see that there is a significant difference in richness between rocks and sand and rocks and eelgrass in both spring and autumn and that those differences survive correction for multiple testing. Richness is still not significantly different between eelgrass and sand.


Next up we are going to test if there are significant differences in ASV richness across substrates i.e. water and sediment.

```{r Substrate ANOVA test }

res_aov_rich_subst <- aov(rich ~ Substrate,
  data = dat)
summary(res_aov_rich_subst)
```

There is a significant difference in richness between sediment and water!
Lets check the residuals for normality.

```{r Substrate ANOVA residuals, message=FALSE}

hist(res_aov_rich_subst$residuals)
qqPlot(res_aov_rich_subst$residuals, id = FALSE)
```

Plotting the residuals from the anova confirms that the residuals are (almost) normally distributed.

Now we'll illustrate these differences on a boxplot across seasons:

```{r grouped boxplot w. substrate sig diff, fig.width = 10, fig.height = 7}

boxplot_substr <- ggplot(dat, aes(x=Substrate, y=rich, fill = habitat)) + 
    geom_boxplot() +
  scale_fill_manual(values = c("yellowgreen", "cornflowerblue","thistle3"),
                     name = "", breaks = c("eelgrass","rocks", "sand"),
                     labels = c("eelgrass","rocks", "sand")) +
  xlab("") + ylab("MOTU richness")+
  theme(legend.position = "right" , text = element_text(size = 18))+
  geom_text_repel(aes(label=cluster),na.rm=TRUE,nudge_y=0.05)+
  geom_pwc(aes(group = Substrate),
           tip.length = 0,
           method = "t_test")

ggadjust_pvalue(
  boxplot_substr, p.adjust.method = "bonferroni",
  label = "{p.adj.signif}") #change to {p.adj.format} to get actual values
```

Now we'll illustrate these differences on a boxplot grouped by season

```{r grouped boxplot w. substrate sig diff w. season, fig.width = 10, fig.height = 7}

boxplot_substr_season <- ggplot(dat, aes(x=Substrate, y=rich, fill = habitat)) + 
    geom_boxplot() +
  scale_fill_manual(values = c("yellowgreen", "cornflowerblue","thistle3"),
                     name = "", breaks = c("eelgrass","rocks", "sand"),
                     labels = c("eelgrass","rocks", "sand")) +
  xlab("") + ylab("MOTU richness")+
  facet_grid(. ~ season)+ 
  theme(legend.position = "top" , text = element_text(size = 18))+
  geom_text_repel(aes(label=cluster),na.rm=TRUE,nudge_y=0.05)+
  geom_pwc(aes(group = Substrate),
           tip.length = 0,
           method = "t_test")

boxplot_substr_adj <- ggadjust_pvalue(
  boxplot_substr_season, p.adjust.method = "bonferroni",
  label = "{p.adj.signif}") #change to {p.adj.format} to get actual values

boxplot_substr_adj

saveRDS(boxplot_substr_adj,"../../Plots/Boxplots/18S_substrate.rds")
```
Looks like any significant differences in 18S ASV richness between water and sediment disappears in the autumn. Lets formally test this by testing the interaction term:

```{r Substrate season interaction ANOVA test }

res_aov_rich_subst_inter <- aov(rich ~ Substrate*season,
  data = dat)
summary(res_aov_rich_subst_inter)
```

There is indeed a significant interaction effect of substrate and season and a main effect of substrate, but no main effect of just season.
We can use the Tukey HSD post-hoc test to dive into the specific comparisons:

```{r Tukey HSD test interaction,}

TukeyHSD(res_aov_rich_subst_inter)
```
We can see that there is no significant difference in 18S richness between water and sediment in the autumn but there is a significant difference in 18S richness between water and sediment in the spring. We can also see that there is a significant difference in 18S richness between water in spring and water in autumn, but not between sediment in spring and sediment in autumn.
From these results it looks like the significant difference in 18S richness between sediment and water is driven by a decrease in richness in water in the autumn.

```{r Venn function}

# Function for plotting euler diagrams with ggplot2 (adapted from: https://gist.github.com/danlooo/d23d8bcf8856c7dd8e86266097404ded)

ggeulerr <- function(combinations, show_quantities = TRUE, show_labels = FALSE, ...) {
  data <-
    eulerr::euler(combinations = combinations) %>%
    plot(quantities = show_quantities) %>%
    pluck("data")
  
  tibble() %>%
    ggplot() +
    ggforce::geom_ellipse(
      data = data$ellipses %>% as_tibble(rownames = "Set"),
      mapping = aes(x0 = h, y0 = k, a = a, b = b, angle = 0, fill = Set),
      alpha = 0.5
    ) +
    geom_text(
      data = {
        data$centers %>%
          mutate(
            label = labels %>% map2(quantities, ~ {
              if (!is.na(.x) && !is.na(.y) && show_labels) {
                paste0(.x, "\n", sprintf(.y, fmt = "%.3g"))
              } else if (!is.na(.x) && show_labels) {
                .x
              } else if (!is.na(.y)) {
                .y
              } else {
                ""
              }
            })
          )
      },
      mapping = aes(x = x, y = y, label = label), fontface="bold", size=2
    ) +
    theme(panel.grid = element_blank()) +
    coord_fixed()
}
```

```{r venn both}

# Subset to clusters where all three habitats were sampled
rep_agg$sc<-metadata$sc[match(row.names(rep_agg),metadata$sshc)]
length(unique(rep_agg$sc)) # 62
hab_reps<-rep_agg %>% group_by(sc) %>% dplyr::summarise(n=n())
three_hreps<-filter(hab_reps,n=="6")
nrow(three_hreps) # 28
both_3hab<-rep_agg[rep_agg$sc %in% three_hreps$sc, ]

# Subset to each habitat
both_3hab$habitat<-metadata$habitat[match(row.names(both_3hab),metadata$sshc)]

eel<-both_3hab[both_3hab$habitat=="eelgrass",]
rock<-both_3hab[both_3hab$habitat=="rocks",]
sand<-both_3hab[both_3hab$habitat=="sand",]

# Remove unnecessary columns
eel<-within(eel,rm(habitat,sc,Substrate))
rock<-within(rock,rm(habitat,sc,Substrate))
sand<-within(sand,rm(habitat,sc,Substrate))

##Convert to presence/absence
eel[eel>0]<-1
rock[rock>0]<-1
sand[sand>0]<-1

# Make species list for each habitat
eel<-eel[,colSums(eel)>0]
eel_sp<-colnames(eel)

rock<-rock[,colSums(rock)>0]
rock_sp<-colnames(rock)

sand<-sand[,colSums(sand)>0]
sand_sp<-colnames(sand)

both_list <- list(Eelgrass = eel_sp,Rock = rock_sp,Sand = sand_sp)

venn_both <- ggeulerr(both_list[1:3]) +
  theme_bw() + # For white background
  theme(axis.title = element_blank(), axis.text = element_blank(), axis.ticks = element_blank()) +  # To remove axis text, titles and ticks
  theme(panel.grid = element_blank()) + # To remove gridlines
  theme(panel.border = element_blank()) + # To remove plot border
  scale_fill_manual(values = c(Eelgrass = "yellowgreen", Sand = "thistle3", Rock = "cornflowerblue"))  #Setting colors for categories
  #+ guides(fill="none") # To remove legend for categories (these are shown with labels anyways)

venn_both

saveRDS(venn_both,"../../Plots/Venn/18S_both.rds")

rep_agg<-within(rep_agg,rm(sc))
```

```{r venn sed}

# Subset to clusters where all three habitats were sampled
sed$sc<-metadata$sc[match(row.names(sed),metadata$sshc)]
length(unique(sed$sc)) # 62
hab_reps<-sed %>% group_by(sc) %>% dplyr::summarise(n=n())
three_hreps<-filter(hab_reps,n=="3")
nrow(three_hreps) # 32
sed_3hab<-sed[sed$sc %in% three_hreps$sc, ]

# Subset to each habitat
sed_3hab$habitat<-metadata$habitat[match(row.names(sed_3hab),metadata$sshc)]

eel<-sed_3hab[sed_3hab$habitat=="eelgrass",]
rock<-sed_3hab[sed_3hab$habitat=="rocks",]
sand<-sed_3hab[sed_3hab$habitat=="sand",]

eel<-within(eel,rm(habitat,sc))
rock<-within(rock,rm(habitat,sc))
sand<-within(sand,rm(habitat,sc))

# Make species list for each habitat
eel<-eel[,colSums(eel)>0]
eel_sp<-colnames(eel)

rock<-rock[,colSums(rock)>0]
rock_sp<-colnames(rock)

sand<-sand[,colSums(sand)>0]
sand_sp<-colnames(sand)

sed_list <- list(Eelgrass = eel_sp,Rock = rock_sp,Sand = sand_sp)

venn_sed <- ggeulerr(sed_list[1:3]) +
  theme_bw(base_size = 14) + # For white background
  theme(axis.title = element_blank(), axis.text = element_blank(), axis.ticks = element_blank()) +  # To remove axis text, titles and ticks
  theme(panel.grid = element_blank()) + # To remove gridlines
  theme(panel.border = element_blank()) + # To remove plot border
  scale_fill_manual(values = c(Eelgrass = "yellowgreen", Sand = "thistle3", Rock = "cornflowerblue"))  #Setting colors for categories
  #+ guides(fill="none") # To remove legend for categories (these are shown with labels anyways)

venn_sed

saveRDS(venn_sed,"../../Plots/Venn/18S_sed.rds")

sed<-within(sed,rm(sc))
```

```{r venn wat}

# Subset to clusters where all three habitats were sampled
wat$sc<-metadata$sc[match(row.names(wat),metadata$sshc)]
length(unique(wat$sc)) # 63
hab_reps<-wat %>% group_by(sc) %>% dplyr::summarise(n=n())
three_hreps<-filter(hab_reps,n=="3")
nrow(three_hreps) # 32
wat_3hab<-wat[wat$sc %in% three_hreps$sc, ]

# Subset to each habitat
wat_3hab$habitat<-metadata$habitat[match(row.names(wat_3hab),metadata$sshc)]

eel<-wat_3hab[wat_3hab$habitat=="eelgrass",]
rock<-wat_3hab[wat_3hab$habitat=="rocks",]
sand<-wat_3hab[wat_3hab$habitat=="sand",]

eel<-within(eel,rm(habitat,sc))
rock<-within(rock,rm(habitat,sc))
sand<-within(sand,rm(habitat,sc))

# Make OTU list for each habitat
eel<-eel[,colSums(eel)>0]
eel_sp<-colnames(eel)

rock<-rock[,colSums(rock)>0]
rock_sp<-colnames(rock)

sand<-sand[,colSums(sand)>0]
sand_sp<-colnames(sand)

wat_list <- list(Eelgrass = eel_sp,Rock = rock_sp,Sand = sand_sp)

venn_wat <- ggeulerr(wat_list[1:3]) +
  theme_bw(base_size = 14) + # For white background
  theme(axis.title = element_blank(), axis.text = element_blank(), axis.ticks = element_blank()) +  # To remove axis text, titles and ticks
  theme(panel.grid = element_blank()) + # To remove gridlines
  theme(panel.border = element_blank()) + # To remove plot border
  scale_fill_manual(values = c(Eelgrass = "yellowgreen", Sand = "thistle3", Rock = "cornflowerblue"))  #Setting colors for categories
  #+ guides(fill="none") # To remove legend for categories (these are shown with labels anyways)

venn_wat

ggsave(filename='../../Plots/Venn/18S_wat.png',venn_sed)

saveRDS(venn_wat,"../../Plots/Venn/18S_wat.rds")

wat<-within(wat,rm(sc))
```

```{r nmds sed habitat}

  #NB., the number of points n should be n > 2k + 1, and preferably higher in global non-metric MDS, and still higher in local NMDS. A k higher than 5 makes it difficult to interpret results

## Parameter settings inspired by https://rpubs.com/CPEL/NMDS

  set.seed(123)
  allmds <- metaMDS(sed, distance = "jaccard", autotransform = FALSE, maxit=999, k=3, trymax=250)

  ## Stress: 0.12
  
  site_scrs <- as.data.frame(scores(allmds, display = "sites"))

  site_scrs$habitat <- metadata$habitat[match(row.names(site_scrs), metadata$sshc)]
  site_scrs$season <- metadata$season[match(row.names(site_scrs), metadata$sshc)]
  
  cent <- aggregate(cbind(NMDS1, NMDS2) ~ habitat, data=site_scrs, FUN = mean)
  segs <- merge(site_scrs, setNames(cent, c("habitat", "oNMDS1", "oNMDS2")), by = "habitat", sort=FALSE)

  plots_hab<-list()
  
  cols <- c("sand"="thistle3","eelgrass"="yellowgreen","rocks"="cornflowerblue")

  nmds_sed<-ggplot(site_scrs, aes(x=NMDS1, y=NMDS2, colour = habitat, shape = season)) + 
  scale_color_manual(values = cols,
                     name = "", breaks = unique(metadata$habitat),
                     labels = unique(metadata$habitat)) +
  geom_point(alpha=1, size=2) +
  coord_fixed() +
  theme_classic() + 
  theme(legend.position = "top", legend.text = element_text(size = 10), legend.title = element_blank(), axis.text = element_text(size = 8), axis.title.x = element_blank(), axis.title.y = element_blank(),plot.margin = unit(c(0,0.4,0,0.4), 'lines'))+
  ggtitle("") +
  #set breaks on x-axis
scale_x_continuous(breaks = c(-1.0,0,1.0))+
  #set breaks on y-axis
scale_y_continuous(breaks = c(-1.5,-1,-0.5,0,0.5,1,1.5,2))

saveRDS(nmds_sed,"../../Plots/NMDS/NMDS_18S_hab_sed.rds")
```

##PERMANOVA analysis
```{r permanova sed}

dist<-vegdist(sed,method="jaccard")

#Make new metadata table for sediment
meta<-as.data.frame(row.names(sed))
row.names(meta)<-meta$`row.names(sed)`
meta$habitat<-metadata$habitat[match(row.names(meta),metadata$sshc)]
meta<-within(meta,rm(`row.names(sed)`))

#Test for homogeneity of multivariate dispersions
b<-betadisper(dist,meta$habitat)
permutest(b)

## p=0.003, so not homogeneous

#Perform a PERMANOVA test
permanova<-adonis2(formula = dist ~ habitat, data = meta)
permanova

##  p=0.001 so significant effect of habitat. R2=0.06

## Same test for season

#Test for homogeneity of multivariate dispersions
meta$season<-metadata$season[match(row.names(meta),metadata$sshc)]
b<-betadisper(dist,meta$season)
permutest(b)

## p=0.026 so not homogeneous

#Perform a PERMANOVA test
permanova<-adonis2(formula = dist ~ season, data = meta)
permanova

## p=0.001 so significant effect of season. R2=0.017
```

```{r nmds wat habitat}

  set.seed(123)
  allmds_w <- metaMDS(wat, distance = "jaccard", autotransform = FALSE, maxit=999, k=3, trymax=250)

  ## Stress: 0.14
  
  site_scrs_w <- as.data.frame(scores(allmds_w, display = "sites"))

  site_scrs_w$habitat <- metadata$habitat[match(row.names(site_scrs_w), metadata$sshc)]
  site_scrs_w$season <- metadata$season[match(row.names(site_scrs_w), metadata$sshc)]
  
  cent <- aggregate(cbind(NMDS1, NMDS2) ~ habitat, data=site_scrs_w, FUN = mean)
  segs <- merge(site_scrs_w, setNames(cent, c("habitat", "oNMDS1", "oNMDS2")), by = "habitat", sort=FALSE)

  plots_hab_w<-list()
  
  nmds_wat<-ggplot(site_scrs_w, aes(x=NMDS1, y=NMDS2, colour = habitat, shape = season)) + 
  scale_color_manual(values = cols,
                     name = "", breaks = unique(metadata$habitat),
                     labels = unique(metadata$habitat)) +
  geom_point(alpha=1,size=2) +
  coord_fixed() +
  theme_classic() + 
  theme(legend.position = "right", legend.text = element_text(size = 10), legend.title = element_blank(), axis.text = element_text(size = 8), axis.title.x = element_blank(), axis.title.y = element_blank(),plot.margin = unit(c(0,0.4,0,0.4), 'lines'))+
  ggtitle("")+
  #set breaks on x-axis
scale_x_continuous(breaks = c(-1.0,0,1.0))+
  #set breaks on y-axis
scale_y_continuous(breaks = c(-1.5,-1,-0.5,0,0.5,1,1.5,2))

saveRDS(nmds_wat,"../../Plots/NMDS/NMDS_18S_hab_wat.rds")
```

##PERMANOVA analysis
```{r permanova wat}

dist<-vegdist(wat,method="jaccard")

#Make new metadata table for water
meta<-as.data.frame(row.names(wat))
row.names(meta)<-meta$`row.names(wat)`
meta$habitat<-metadata$habitat[match(row.names(meta),metadata$sshc)]
meta<-within(meta,rm(`row.names(wat)`))

#Test for homogeneity of multivariate dispersions
b<-betadisper(dist,meta$habitat)
permutest(b)

## p=0.04 so not homogeneous

#Perform a PERMANOVA test
permanova<-adonis2(formula = dist ~ habitat, data = meta)
permanova

## p=0.001, and R2=0.03

## Same test for season

#Test for homogeneity of multivariate dispersions
meta$season<-metadata$season[match(row.names(meta),metadata$sshc)]
b<-betadisper(dist,meta$season)
permutest(b)

## p=0.143, so homogeneous

#Perform a PERMANOVA test
permanova<-adonis2(formula = dist ~ season, data = meta)
permanova

## p=0.001 and R2=0.05
```

```{r nmds sed salinity}

## Get salinities
coord<-read.table("../../Tekstfiler/Across_barcodes/merged_metadata_230427.txt", sep="\t", row.names=1, header=T, check.names=FALSE)
metadata$new_name<-paste(metadata$season,metadata$cluster,metadata$habitat,sep="_")
metadata$salinity <- coord$Salinity[match(metadata$new_name,row.names(coord))]

site_scrs$salinity <- metadata$salinity[match(row.names(site_scrs), metadata$sshc)]

cent <- aggregate(cbind(NMDS1, NMDS2) ~ salinity, data=site_scrs, FUN = mean)
segs <- merge(site_scrs, setNames(cent, c("salinity", "oNMDS1", "oNMDS2")), by = "salinity", sort=FALSE)

# set breaks and colours
breaks <- c(7,11,17,22,27,33)
cols <- RColorBrewer::brewer.pal(9, "RdYlBu")

plots<-list()

  nmds_sal<-ggplot(site_scrs, aes(x=NMDS1, y=NMDS2)) + 
  scale_colour_stepsn(colours = rev(cols),
                    breaks = breaks,
                    name = "Salinity") +
  geom_point(alpha=1, size=2, aes(colour = salinity, shape = season)) +
  coord_fixed() +
  theme_classic() + 
  theme(legend.position = "right", legend.text = element_text(size = 10), legend.title = element_text(size = 12), axis.text = element_text(size = 8), axis.title.x = element_blank(), axis.title.y = element_blank(),plot.margin = unit(c(0,0.4,0,0.4), 'lines'))+
  ggtitle("")+
  #set breaks on x-axis
scale_x_continuous(breaks = c(-1.0,0,1.0))+
  #set breaks on y-axis
scale_y_continuous(breaks = c(-1.5,-1,-0.5,0,0.5,1,1.5,2))
  
saveRDS(nmds_sal,"../../Plots/NMDS/NMDS_18S_sal_sed.rds")
```

```{r nmds wat salinity}

## Get salinities
site_scrs_w$salinity <- metadata$salinity[match(row.names(site_scrs_w), metadata$sshc)]

cent <- aggregate(cbind(NMDS1, NMDS2) ~ salinity, data=site_scrs_w, FUN = mean)
segs <- merge(site_scrs_w, setNames(cent, c("salinity", "oNMDS1", "oNMDS2")), by = "salinity", sort=FALSE)

# set breaks and colours
breaks <- c(7,11,17,22,27,33)
cols <- RColorBrewer::brewer.pal(9, "RdYlBu")

plots_wat<-list()

  nmds_sal_w<-ggplot(site_scrs_w, aes(x=NMDS1, y=NMDS2)) + 
  scale_colour_stepsn(colours = rev(cols),
                    breaks = breaks,
                    name = "Salinity") +
  geom_point(alpha=1, size=2, aes(colour = salinity, shape = season)) +
  coord_fixed() +
  theme_classic() + 
  theme(legend.position = "right", legend.text = element_text(size = 10), legend.title = element_text(size = 12), axis.text = element_text(size = 8), axis.title.x = element_blank(), axis.title.y = element_blank(),plot.margin = unit(c(0,0.4,0,0.4), 'lines'))+
  ggtitle("")+
  #set breaks on x-axis
scale_x_continuous(breaks = c(-1.0,0,1.0))+
  #set breaks on y-axis
scale_y_continuous(breaks = c(-1.5,-1,-0.5,0,0.5,1,1.5,2))
  
saveRDS(nmds_sal_w,"../../Plots/NMDS/NMDS_18S_sal_wat.rds")
```

```{r nmds both}

##Convert to presence/absence
both<-within(rep_agg,rm(Substrate))
both[both>0]<-1

#NB., the number of points n should be n > 2k + 1, and preferably higher in global non-metric MDS, and still higher in local NMDS. A k higher than 5 makes it difficult to interpret results

## Parameter settings inspired by https://rpubs.com/CPEL/NMDS

  set.seed(123)
  allmds <- metaMDS(both, distance = "jaccard", autotransform = FALSE, maxit=999, k=3, trymax=250)

  ## Stress: 0.12
  
  site_scrs <- as.data.frame(scores(allmds, display = "sites"))
  site_scrs$Substrate <- metadata$substrate_type[match(row.names(site_scrs), metadata$sshc)]

  cent <- aggregate(cbind(NMDS1, NMDS2) ~ Substrate, data=site_scrs, FUN = mean)
  segs <- merge(site_scrs, setNames(cent, c("Substrate", "oNMDS1", "oNMDS2")), by = "Substrate", sort=FALSE)

  cols <- c("sediment" = "#d8b365", "water" = "#5ab4ac")

  nmds_both<-ggplot(site_scrs, aes(x=NMDS1, y=NMDS2, colour = Substrate)) + 
  scale_color_manual(values = cols,
                     name = "", breaks = unique(metadata$substrate_type),
                     labels = unique(metadata$substrate_type)) +
  geom_point(alpha=1, size=2) +
  coord_fixed() +
  theme_classic() + 
  theme(legend.position = "right", legend.text = element_text(size = 10), legend.title = element_text(size = 8), axis.text = element_text(size = 8), axis.title.x = element_blank(),axis.title.y = element_text(size = 8),plot.margin = unit(c(0,0.4,0,0.4), 'lines'))+
  ggtitle("")+
  #set breaks on x-axis
scale_x_continuous(breaks = c(-1.0,0,1.0))+
  #set breaks on y-axis
scale_y_continuous(breaks = c(-1.5,-1,-0.5,0,0.5,1,1.5,2))

saveRDS(nmds_both,"../../Plots/NMDS/NMDS_18S_both.rds")
```